Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-03-19 manuscript.qmd
add table with param estimates

---
title: "Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer"
authors:
  - name: Roman Ludwig
    orcid: 0000-0001-9434-328X
    email: roman.ludwig@usz.ch
  - name: Yoel Perez Haas
    email: yoel.perezhaas@usz.ch
  - name: Jan Unkelbach
    orcid: 0000-0002-4275-990X
    email: jan.unkelbach@usz.ch
format:
  html:
    code-fold: true
  pdf:
    keep-tex: true
    include-in-header:
    - text: |
        \usepackage{multirow}
        \usepackage{centernot}
jupyter: python3
execute: 
  cache: true
crossref:
  fig-prefix: "figure"
  tbl-prefix: "table"
  eq-prefix: "equation"
  sec-prefix: "section"
---

@@

data = pd.read_csv(paths.filtered, header=[0,1,2])
data[COL.t_stage] = data[COL.t_stage].apply(lambda x: "early" if x <= 2 else "late")
grouped = data.groupby(by=[COL.midext, COL.t_stage, COL.ipsi_III])

@@

involved = num_involved.join(
  100 * percent_involved,
  rsuffix=" (%)",
).sort_index(axis="columns")
involved.columns = pd.MultiIndex.from_product([["I", "II", "III", "IV"], ["n", "%"]])
(
  involved.style
  .format(precision=2)
  .apply(right_align, axis="index")
)
```

@@

```{python}
#| echo: false
#| label: fig-burnin-history
#| layout-ncol: 1
#| layout-nrow: 1
#| fig-cap: Monitoring quantities during the burn-in phase of the parameter sampling.
#| fig-subcap: 
#|   - "Autocorrelation time"
#|   - "log likelihood"
from collections import namedtuple

@@

>>>>

---
title: "Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer"
authors:
  - name: Roman Ludwig
    orcid: 0000-0001-9434-328X
    email: roman.ludwig@usz.ch
  - name: Yoel Perez Haas
    email: yoel.perezhaas@usz.ch
  - name: Jan Unkelbach
    orcid: 0000-0002-4275-990X
    email: jan.unkelbach@usz.ch
format:
  html:
    code-fold: true
  pdf:
    keep-tex: true
    include-in-header:
    - text: |
        \usepackage{multirow}
        \usepackage{centernot}
jupyter: python3
execute: 
  cache: false
crossref:
  fig-prefix: "figure"
  tbl-prefix: "table"
  eq-prefix: "equation"
  sec-prefix: "section"
---

@@

data = pd.read_csv(paths.filtered, header=[0,1,2])
data[COL.t_stage] = data[COL.t_stage].apply(lambda x: "early" if x <= 2 else "late")
early_midext_data = (data.loc[data[COL.t_stage] == "early", COL.midext] == True).mean()
late_midext_data = (data.loc[data[COL.t_stage] == "late", COL.midext] == True).mean()
grouped = data.groupby(by=[COL.midext, COL.t_stage, COL.ipsi_III])

@@

involved = num_involved.join(
  100 * percent_involved,
  rsuffix=" (%)",
).sort_index(axis="columns")
involved.columns = pd.MultiIndex.from_product(
  [["I", "II", "III", "IV"],
   ["n", "%"]],
  names=["LNL", ""]
)
(
  involved.style
  .format(precision=2)
  .apply(right_align, axis="index")
)
```

@@

```{python}
#| echo: false
#| label: fig-burnin-history
#| layout-ncol: 1
#| layout-nrow: 1
#| fig-cap: Monitoring quantities during the burn-in phase of the parameter sampling.
from collections import namedtuple

@@

## Parameter Estimates

```{python}
#| label: tbl-midline-params
#| tbl-cap: Mean sampled parameter estimates of the midline model and the respective standard deviation.
#| echo: false

import yaml

import pandas as pd
from emcee.backends import HDFBackend
from lyscripts.utils import create_model

with open("models/midline/params.yaml", mode="r") as file:
  config = yaml.safe_load(file)

model = create_model(config)
backend = HDFBackend("models/midline/samples.hdf5", read_only=True)
chain = backend.get_chain(flat=True)

param_names = list(model.get_params().keys())
param_means = chain.mean(axis=0)
param_stds = chain.std(axis=0)
model.set_params(*param_means)
early_midext_prob = model.get_distribution("early").pmf @ model.midext_evo()[:,1]
late_midext_prob = model.get_distribution("late").pmf @ model.midext_evo()[:,1]

params_table = pd.DataFrame({
  "name": param_names,
  "mean": param_means,
  "std": param_stds,
})
(
  params_table.style
  .format("{:.2%}", subset=["mean", "std"])
  .apply(right_align, subset=["mean", "std"])
  .hide()
)
```

The predicted prevalence of midline extension for early T-category tumors is `{python} f"{early_midext_prob:.2%}"`. For a late T-category tumor, it is `{python} f"{late_midext_prob:.2%}"`.

In the data, the prevalence of midline extension for early T-category tumors is `{python} f"{early_midext_data:.2%}"`. For late T-category tumors it is `{python} f"{late_midext_data:.2%}"`.

- Discrepancy between data and model is due to late T-stage's time prior
- still tied to other parameters
- cannot fully represent spread of midline extension and also represent spread params
